The Supervisionary proof-checking kernel

### 2021-10-23 paper/prisc-2022.tex
prisc-2022: from the ashes a fire shall be woken

@@

>>>>

Interactive theorem proving software is typically designed around a proof-checking \emph{kernel}.
Proofs are constructed by submitting theorem derivations to the kernel for authentication, and then composing them with other authentic theorems.
A proof may be large and complex, as may be the software components, both interactive and automated, used to construct it.
However, the small and simple kernel is soley responsible for validating it.
This implementation strategy---known as the ``LCF-style'', after Milner's system that first implemented it---is a reliability mechanism, aiming to ensure that any purported theorem produced by the system is indeed a theorem.

In an operating system, the \emph{kernel} is the most privileged component, responsible for mediating interaction by less privileged components with the hardware and with each other.
An unpriviledged process interacts with the rest of the system by issuing \emph{system calls} accross a hardware privilege boundary to the operating system kernel, which may contain requests for hardware resources, accesses to shared system services such as the filesystem, or messages addressed to other processes.
In this way, the operating system kernel \emph{supervises} less priviledged processes.

In a typical proof assistant, the boundary between a proof-checking kernel and the rest of the system is a programming-language level module boundary.
In this work, we explore the implications of a proof-checking kernel which acts as a supervisor to the rest of theorem proving system, across a more concrete privilege boundary.

\paragraph{The Supervisionary system interface}

\emph{Supervisionary} is the name of our prototype implementation of a proof-checking system for Gordon's HOL structured as supervisory software, for expressing and experimenting with the ideas presented in this paper.
In particular, our prototype specifies a \emph{system interface} to the proof-checking kernel which serve as the substance of communication across any of the concrete privilege boundries described in the body of this paper.

% this bit is copied mostly verbatim from the previous version

The kernel manages various private heaps within which \emph{kernel objects} are allocated, corresponding to the paraphenalia of any HOL implementation---type-formers, types, term constants, terms, and theorems---and never directly exposed to untrusted code.
Kernel objects are allocated in response to system calls like:
\begin{displaymath}
\mathtt{Term.Handle.AllocateApplication(left, right, out)}
\end{displaymath}
Here, both $\mathtt{left}$ and $\mathtt{right}$ are \emph{kernel handles}, assumed to point-to allocated terms, whilst $\mathtt{out}$ points-to a buffer in untrusted code's memory.
If neither $\mathtt{left}$ nor $\mathtt{right}$ \emph{dangle}, and the types of their referents match, a fresh handle is generated which points-to a new HOL term application object, with internal pointers to the functional- and argument-terms.
This handle is returned to the caller via the $\mathtt{out}$ pointer.

The manipulation and querying of kernel objects is performed defensively by the kernel itself on behalf of untrusted code.
The kernel is careful to maintain invariants such as the \emph{inductivity} of its heaps, with nodes in the kernel object graph pointing-to allocated objects at all times.

Space constraints prevent us from describing the entire Supervisionary ABI for working with, and on, kernel objects.
However, note that theorems are also constructed in a similar way to terms, inductively building derivation trees.
For example, the HOL symmetry rule is exposed as:
\begin{displaymath}
\mathtt{Theorem.Handle.AllocateSym(pre, out)}
\end{displaymath}
Here, $\mathtt{pre}$ points-to an existing theorem object $\Gamma \vdash r = s$ and after succeeding, passing obvious checks, $\mathtt{out}$ contains a handle that points-to a newly-allocated theorem $\Gamma \vdash s = r$.

\paragraph{More flexible theorem proving tools}

Immediately, we can re-imagine the implementation of the foundations of familiar theorem proving tools.
For this case, the unpriviledged bulk of the theorem proving system runs inside of a WebAssembly (Wasm) sandbox, supervised by the proof-checking kernel which runs as native code in the process hosting the sandbox.
In the case of a typical LCF-style proof assistant such as HOL4, the entire system could be compiled to Wasm, and the module implementing the kernel could be modified to act as a proxy in front of the Supervisionary kernel outside of the sandbox.
This modification has immediate implications for the reliability and auditability of the system, because the only critical component now isolated from the bulk of the system by a priviledged boundary.

% versus just type system
% even memory corruption vulns in bulk aren't a problem
% next step: polyglot theorem proving software
% another benefit: strong isolation boundaries within your proof: isolate important spec from the large proof

\paragraph{Generalized runtime verification}

% I can draft this section on Saturday.

\paragraph{More}

TODO

% VM/PA

% more about our prototype?

\iffalse

@@

\fi
